# Coursera Reproducable Research Assignment 1


### Loading and Processing the data

Loading the ggplot, lattice, and plyr packages.  These packages will be used
throughout the analysis.
``` {r echo = TRUE}
library(ggplot2)
library(plyr)
library(dplyr)
library(lattice)
```

Setting the working directory then downloading the dataset.

``` {r echo = TRUE}
setwd("C:/Users/John/Documents/Coursera/reproducibleresearch/project1")
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
path <- getwd()
download.file(url,destfile=file.path(path, "Activity monitoring data.zip"))
unzip(zipfile = "Activity monitoring data.zip")
activitydata <-read.csv("activity.csv")
```
###What is mean total number of steps taken per day?

Aggregating the dataset in order to find the average number of steps per day.
```{r echo = TRUE}
avgTable <- aggregate(activitydata$steps ~ activitydata$date, FUN=sum, )
colnames(avgTable)<- c("Date", "Steps")
```

Creating a histogram of the average number of steps per day.
```{r echo = TRUE}
hist(avgTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```

Displaying mean and median summary statistics.
```{r echo = TRUE}
as.integer(mean(avgTable$Steps))
as.integer(median(avgTable$Steps))
```

### What is the average daily activity pattern?

Using previously stored data, remove any NA values and create a clean 
table for the average number of steps.
```{r echo = TRUE}
clean <- activitydata[!is.na(activitydata$steps),]
cleanTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))
```

Create a ggplot line graph (best way to plot time-series data) to display
number of average steps per interval.
``` {r echo = TRUE}
p <- ggplot(cleanTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```

Maximum number of steps per interval:
``` {r echo = TRUE}
maxSteps <- max(cleanTable$Avg)
```
Which interval contains the maximum average number of steps?
``` {r echo = TRUE}
cleanTable[cleanTable$Avg==maxSteps,1]
```

### Imputing Missing Values

Calculate and report the total number of missing values in the dataset.
``` {r echo = TRUE}
nrow(activitydata[is.na(activitydata$steps),])
```

My strategy for NA handling was to remove the NA values and replace them with mean values.
``` {r echo = TRUE}
# find positions of NA values
napos <- which(is.na(activitydata$steps))
# create a vector with mean values
meanvec <- rep(mean(activitydata$steps, na.rm=TRUE), times=length(napos))
# replace the NA values with the means
activitydata[napos, "steps"] <- meanvec
rm(mean_vec, na_pos)
# calculate the total number of steps each day (NA values removed)
sumdata <- aggregate(activitydata$steps, by=list(activitydata$date), FUN=sum)
# Rename the variables
colnames(sumdata) <- c("date", "total")
```

Plotting a histogram with the total number of steps taken each day.
``` {r echo = TRUE}
hist(sumdata$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")
```

Calculate summary statistics with NA values imputted
``` {r echo = TRUE}
mean(sumdata$total)
median(sumdata$total)
```

### Are there differences in activity patterns between weekdays and weekends?

The Lubridate has a function that will allow us to easily convert a date to
the corresponding day type (monday, tuesday, etc.).  This is similar
to a date dimension table in an a data warehouse.
``` {R echo = TRUE}
# loading lubridate package to prase day categories based on dates
library(lubridate)
```

Here, we will parse the date to find if it's on a weekday or weekend
based on the date.  Then we'll summarize the data into a table
and plot a chart.
``` {r echo = TRUE}
# Create new category based on the days of the week
sumdata$DayCategory <- ifelse(wday(mdy_hms(sumdata$date), label = TRUE) %in% c("Sat", "Sun"), "weekend", "weekday")

## Summarize data by interval and type of day
catasum <- ddply(sumdata, .(interval, DayCategory), summarize, Avg = mean(total))

##Plot data in a panel plot
#xyplot(Avg~interval|DayCategory, data=catsum, type="l",  layout = c(1,2),
#       main="Average Steps per Interval Based on Type of Day", 
#       ylab="Average Number of Steps", xlab="Interval")
```





